Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-07 biology-data.tex
write most of intro
edit backcalc methods
Resolve to dos

In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS), first introduced in\todo{ref where introducing studies}.
Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
Therefore, the results from these studies are more generalisable to the entire UK population.

@@

In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (inclusive).
\todo{choice of period might be better in intro}
This period starts when CIS was expanded, and ends when vaccination crossed 10\% of England's population.
Vaccination complicates the dynamics of transmission and viral load.
Analysis of data including vaccinated individuals requires more complex models, with specific adaptations to SARS-CoV-2.
These are beyond the scope of this thesis, which primarily intends to show the potential of a study such as CIS.

During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
Following this, the recruitment then decreases and stabilises, as does the rate of swabs taken (see \cref{biology-data:fig:CIS-recruitment}(B)).
The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
Because the testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.

@@

The main result reported by the study was what they defined as the \emph{positivity rate}, the proportion of people in a given (sub)population that (if they were tested using RT-PCR) would test positive for SARS-CoV-2 at a defined point in time~\autocite{cisMethodsONS}.
A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
Larger time periods (\ie weeks instead of days) allow more granular estimates in other dimensions, most commonly geography.

@@

\subsubsection{Computational limitations and privacy protection}

>>>>

In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
Therefore, the results from these studies are more generalisable to the entire UK population.

@@

In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (see \cref{E-intro:sec:aims}).
During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
Following this, the recruitment then decreases and stabilises, as does the rate of swabs taken (see \cref{biology-data:fig:CIS-recruitment}(B)).
The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
Because the testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.

@@

The main result reported by the study was what they defined as the \emph{positivity rate}, the proportion of people in a given (sub)population that (if they were tested using RT-PCR) would test positive for SARS-CoV-2 at a defined point in time~\autocite{cisMethodsONS}.
A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is negligible

Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
Larger time periods (\ie weeks instead of days) allow more granular estimates in other dimensions, most commonly geography.

@@

\subsubsection{Computational limitations and privacy protection} \label{biology-data:sec:SRS}

### 2024-03-07 cis-perfect-testing.tex
write most of intro
Resolve to dos

\todo[inline]{Are the following paragraphs better here or in the intro chapter when motivating each chapter and the overall narrative of the thesis?}
In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
This choice is motivated by the computational limitations of the SRS and the CIS's study design (see \cref{E-intro:sec:cis}).
The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.

There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging.
The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
The issue with random effects is made worse by the CIS's study design.
Infrequent testing means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.

@@

\subsection{Unobserved episode beginning and end times}

@@

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

>>>>

@@

\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}

@@

Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

### 2024-03-07 incidence-prevalence.tex
Resolve to dos

\chapter{Statistical and epidemiological background} \label{inc-prev}

\todo[inline]{%
  Chapter 3 needs a better title: this is not all the statistical and epidemiological background and some of it is mathematical modelling / disease dynamics.
  Perhaps: ``Data generating procsses and disease dynamics''.
  Would need to define disease dynamics.
}

%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}

@@

\todo[inline]{Does something need to link these sections?}

@@

Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:cis}), and incidence estimates are the objective of this thesis.
This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
However, no estimates of this quantity in the general population exist\todo{ref lit review of duration}.
Therefore, I now turn to estimating the duration of RT-PCR positivity.

>>>>

\chapter{Data generating process} \label{inc-prev}

@@

@@

Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:cis}), and incidence estimates are the objective of this thesis.
This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
However, no estimates of this quantity in the general population exist (see \cref{E-intro:sec:duration-SARS-CoV-2}).
Therefore, I now turn to estimating the duration of RT-PCR positivity.
